Layer dependence as a measure of local dependence

### 2015-04-17 ld8.tex
pdj changes

Remaining sections are structured as follows. Section \aref{sintroduction} defines and analyzes layer dependence. Section \aref{sldcurve} demonstrates the appropriateness of layer dependence as a local dependence measure using several copulas. Section \aref{sdecompose} explains the behaviour of layer dependence by decomposing it into a negative function of discordance and dispersion. Section \aref{scoherence} describes coherence properties of layer dependence. Links to existing literature are highlighted in  \sref{sliterature}. Further properties of layer dependence are described in  \sref{sproperties}. Section \aref{sfitting} applies layer dependence to copula modeling, and uses historical stock returns as an illustration. Section \aref{saltoverall} discuss alternative coherent measures of overall dependence based on weighted averages of layer dependence. Section \aref{sconclusion} concludes.

@@

Suppose $u$ and $v$ are percentile ranks of continuous random variables $x$ and $y$.  Then $(u,v)$ has standard uniform marginals and joint distribution $C$, a copula \citep{nelson1999ic}. Of interest is the dependence structure underlying $(x,y)$.

Spearman's $\rho$ \citep{embrechts2002correlation} between $x$ and $y$  is a common measure of overall rank dependence, the dependence between percentile ranks $u$ and $v$. Rank dependence is not distorted by marginal distributions unlike for example Pearson's correlation \citep{mcneil2005qrm} which is generally not $\pm 1$ when $x$ and $y$ are perfectly dependent unless $x$ and $y$ are linear in one another. Spearman's $\rho$ can also be applied to estimate copula parameters using the method of moments (\cite{kojadinovic2010comparison}, \cite{bouye2000copulas}). However a drawback of Spearman's $\rho$ is that it does not capture local dependence which may vary across the joint distribution as explained in section \aref{sreview}. This drawback is addressed by a generalisation of Spearman's $\rho$ discussed in the following.

Spearman's $\rho$ is defined and can be rewritten as
\begin{equation}\label{spearman}
\rho \equiv \cor(v,u) = \frac{\cov(v,u)}{\cov(u,u)}
=\frac{\int_0^1 \cov\{v,(u>\alpha)\} \de \alpha }{\int_0^1 \cov\{u,(u>\alpha)\} \de \alpha }
\end{equation}
where $\cov$ and $\cor$ represent covariance and correlation, respectively. In addition $(u>\alpha)$ is a random variable equal to 1 if $u>\alpha$ and 0 otherwise. The final expression in \eref{spearman} relies on the ``layer decomposition" of $u$
\begin{equation}\label{decompose}
u= \int_0^1 (u>\alpha)\de\alpha\ .
\end{equation}
The left hand side of \eref{decompose} is $\int_0^u 1\de \alpha=u$. \eref{decompose} decomposes $u$ into layers $(u>\alpha)\de\alpha$ across $0\le\alpha\le 1$. The term layer is used since integrating $(u>\alpha)\de\alpha$ over a subset $[a,b]$ of the unit interval yields $\min\{\max(u-a,0),b-a\}$ which corresponds to layers covered by excess-of-loss reinsurance \citep{wang1995insurance} and debt tranches \citep{mandel2012role}. The infinitesimally small layer $(u>\alpha)\de\alpha$ captures movements in $u$ at $\alpha$ since it jumps when $u$ crosses $\alpha$ and is constant over other values of $u$.

Removing the integrals in the final expression of \eref{spearman} yields the $\alpha$--layer dependence  between $v$ and $u$:
\begin{equation}\label{definition}
\ell_\alpha \equiv \frac{\cov\{v,(u>\alpha)\}}{\cov\{u,(u>\alpha)\}}
=\frac{\cor\{v,(u>\alpha)\}}{\cor\{u,(u>\alpha)\}} \cq 0\leq\alpha\leq 1\;,
\end{equation}
Denominators in \eref{definition} are independent of $C$ and imply $\ell_\alpha=1$ if $u=v$ and $\ell_\alpha=-1$ if $u=1-v$.  Further $-1\leq \ell_\alpha \leq 1$. Independence implies $\ell_\alpha=0$. These and other properties of $\ell_\alpha$ are explored in sections \aref{scoherence} and \aref{sproperties}. Since $(u>\alpha)$ captures movements in $u$ at $\alpha$, $\ell_\alpha$ generalises Spearman's $\rho$ by measuring local dependence between $v$ and $u$ at $\alpha$. This interpretation is further discussed below and in subsequent sections.

The local dependence interpretation of $\ell_\alpha$ can be alternatively shown by expanding the covariances  in \eref{definition} and manipulating to yield
\begin{equation}\label{gapexp}
\ell_\alpha = \frac{\E(v|u>\alpha)-\E(v|u\leq \alpha)}{\E(u|u>\alpha)-\E(u|u\leq \alpha)}
=2 \left\{\E(v|u>\alpha)-\E(v|u\leq \alpha)\right\}\;,
\end{equation}
where $\E$ calculates expectations using $C$. The middle expression in \eref{gapexp} is the expected change in $v$ relative to the expected change in $u$ when $u$ crosses $\alpha$. The latter is $0.5$ for all $\alpha$, yielding the final expression in \eref{gapexp}. Hence large $\ell_\alpha$ implies $v$ is sensitive to movements in $u$ across $\alpha$, indicating strong dependence between $v$ and $u$ at $\alpha$. When $\ell_\alpha=0$, $v$ is unchanged on average when $u$ crosses $\alpha$, and $u$ and $v$ are independent at $\alpha$.

Rewriting Spearman's $\rho$ in \eref{spearman} in terms of $\ell_\alpha$ yields
\begin{equation}\label{wtdaverage}
\rho = \frac{\int_0^1 \cov\{u,(u>\alpha)\} \ell_\alpha \de \alpha }{\int_0^1 \cov\{u,(u>\alpha)\} \de \alpha }
= \int_0^1 w_\alpha \ell_\alpha \de \alpha
\cq w_\alpha\equiv 6\alpha(1-\alpha) \; .
\end{equation}
Hence Spearman's $\rho$ is a weighted average of $\ell_\alpha$ across $0\le\alpha\le 1$. The weights are $w_\alpha$ and integrate to $1$. Therefore whilst Spearman's $\rho$ measures overall dependence between $u$ and $v$, $\ell_\alpha$ performs the same but locally and averaging values of $\ell_\alpha$ yields $\rho$. Further note $w_\alpha$ has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. This implies Spearman's $\rho$ places less emphasis on tail dependence which may be undesirable in finance or insurance where tail dependence is present. Modifying the weights leads to alternate measures of overall dependence. This is further explored in section \aref{saltoverall}.

@@

The nine panels in \fref{fillustration} display $(u,v)$ scatterplots of exchangeable copulas $C(u,v)=C(v,u)$, and their layer dependence curves $\ell_\alpha$ for all $0\le\alpha\le 1$. $\ell_\alpha$ and $\alpha$ are plotted on the same axes as $v$ and $u$, respectively. Each copula has Spearman's $\rho=0.6$. The intention of \fref{fillustration}  is to illustrate that copulas with the same overall dependence exhibit different local dependence structures which are captured by layer dependence.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \resizebox{40mm}{!}{\includegraphics{normal.eps}}
            \resizebox{40mm}{!}{\includegraphics{student.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural2.eps}} \\
      \resizebox{40mm}{!}{\includegraphics{gumbel.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural1.eps}}
      \resizebox{40mm}{!}{\includegraphics{clayton.eps}} \\
            \resizebox{40mm}{!}{\includegraphics{frank.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural3.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural4.eps}} \\
    \end{tabular}
    \caption{Copulas with the same  $\rho_S=0.6$ but different layer dependence curves $\ell_\alpha$ over $0\leq\alpha\le 1$ (red curves) . }
    \label{fillustration}
  \end{center}
\end{figure}

@@

If $(u,v)$ is exchangeable then layer dependence $\ell_\alpha$ measures the lack of discordance and dispersion at $\alpha$:
\begin{equation}\label{decomposition}
\ell_\alpha = 1-2(1+\gamma_\alpha) \delta_\alpha \;,
\end{equation}
where
$$
\gamma_\alpha \equiv \cor\{(u\leq\alpha),(v>\alpha)\} =\cor\{(u>\alpha),(v\leq\alpha)\}\;,
$$
$$
\delta_\alpha \equiv\E\left\{(|u-v|)|(u-\alpha)(v-\alpha)<0\right\}  \; .
$$
A proof of \eref{decomposition} is  below.  The correlation $-1\leq\gamma_\alpha\leq 1$ measures the tendency for $(u,v)$ to be discordant at $\alpha$: opposite signs on $u-\alpha$ and $v-\alpha$.  The expectation $0\leq \delta_\alpha\leq 1$ measures the dispersion between discordant points $u$ and $v$ at $\alpha$.

@@

Note $\cov\{(u>\alpha),(v>\beta)\}=C(\alpha,\beta)-\alpha\beta$ and apply \eref{decompose} to $v$, to derive
\begin{equation}\label{copula}
\ell_\alpha = \frac{\int_0^1 \cov\{(u>\alpha),(v>\beta)\} \de\beta }{\alpha(1-\alpha)/2}
= \frac{2 \int_0^1 C(\alpha,\beta) \de \beta- \alpha}{\alpha(1-\alpha)} \;.
\end{equation}
Thus $\ell_\alpha$ integrates copulas to reduce their dimension from two and one, and scales the result to ensure it lies between $\pm 1$.

>>>>

Remaining sections are  as follows. Section \aref{sintroduction} defines and analyzes layer dependence. Section \aref{sldcurve} demonstrates the appropriateness of layer dependence as a local dependence measure using several copulas. Section \aref{sdecompose} explains the behaviour of layer dependence by decomposing it into a negative function of discordance and dispersion. Section \aref{scoherence} describes coherence properties of layer dependence. Links to existing literature are highlighted in  \sref{sliterature}. Further properties of layer dependence are described in  \sref{sproperties}. Section \aref{sfitting} applies layer dependence to copula modeling, and uses historical stock returns as an illustration. Section \aref{saltoverall} discuss alternative coherent measures of overall dependence based on weighted averages of layer dependence. Section \aref{sconclusion} concludes.

@@

A familiar construct in the study of bivariate dependence is Spearman's correlation  \citep{embrechts2002correlation} which is the ordinary correlation between the  ranks of the two variables.   Spearman's correlation $\rho$ measures the correlation between the ranks of two variables and is  one measure in the arsenal for assessing relationships on the percentile or copula scale.   Rank dependence is not distorted by marginal distributions unlike for example Pearson's correlation \citep{mcneil2005qrm} which measures the degree of linear relationship in the original scale. Spearman's $\rho$ can  be applied to estimate copula parameters using the method of moments (\cite{kojadinovic2010comparison}, \cite{bouye2000copulas}). Spearman's correlation suffers from shortcomings and, as an aggregate measure, may be inappropriate for assessing  local dependence which may vary across the joint distribution including the tails.

A familiar construct in reinsurance is a  loss layer.  For example the 95\%-96\% loss layer of a loss $x$ is that portion of loss $x$ between the 95th and 96th percentile
$$
\min\{(x-x_{0.95})^+,x_{0.96}-x_{0.95}\}\ ,
$$
where $()^+$ indicates the positive part of the expression inside the brackets.  Layers arise in the contex of  excess--of--loss reinsurance \citep{wang1995insurance} and debt tranches \citep{mandel2012role}.

A reinsurer  may be interested in the correlation between a  loss layer from an earthquake disaster and say  earthquake severity as measured on  the Richter scale.   The correlation is an imperfect measure as it is scale dependent and ignores the influence of other variables.   Nevertheless it does provide preliminary guidance.

In this paper we combine the  ideas of rank correlation and correlation between one variable and the loss layer of another.   Thus we consider variables on the percentile scale $u$ and $v$, and consider for example the correlation between say the 95\%-96\% loss layer of $u$ and the percentile $v$.   Since variables are on the  percentile scale, the percentile loss layer reduces to
\begin{equation}\label{eg}
\min\{(u-u_{0.95})^+,u_{0.96}-u_{0.95}\} = \min\{(u-\alpha)^+,\de\alpha\}\ , 
\end{equation}
where in this example $\alpha=0.95$ and $\de \alpha=0.01$.

\subsection{Layers on the percentile scale}
The last expression in \eref{eg} can be approximated by, using  the familiar infinitesimal notation,
\begin{equation}\label{ld}
I_\alpha(u)\de \alpha=\de (u-\alpha)^+  \cq I_\alpha(u) \equiv \left\{\begin{array}{cc}0\ , & u\le\alpha\ ,\\ 1\ ,& u>\alpha\ .\end{array}\right.
\end{equation}
The  left hand side expression of \eref{ld} is called the $\alpha$--layer of $u$ at $\alpha$.  (Graph for $(u-\alpha)^+$?)

As often exploited  in reinsurance, any loss or variable can be thought of as a sum of layers:
\begin{equation}\label{decompose}
u=\int_0^u 1\de \alpha= \int_0^1 I_\alpha(u)\de\alpha \ ,
\end{equation}

\subsection{Layer dependence}

Spearman's correlation is the standardized covariance between $v$ and $u$:
$$
\rho = \frac{\cov(v,u)}{\cov(u,u)}
$$
   This numerator equals
\begin{equation}\label{spearman}
\cov(v,u)=\cov\left\{v,\int_0^1I_\alpha(u)\de \alpha \right\}= \int_0^1 \cov\{v,I_\alpha(u)\} \de \alpha \ .
\end{equation}
Write
\begin{equation}\label{ellalpha}
\ell_\alpha \equiv \frac{\cov\{v,I_\alpha(u)\}}{\cov\{u,I_\alpha(u)\}} 
\end{equation}
(*** Insert something here about the insurance meaning of $\ell_\alpha$.)

then 
\begin{equation}\label{definition}
\rho = \int_0^1 \ell_\alpha\cov\{u,I_\alpha(u)\} \de \alpha =  \int_0^1 \ell_\alpha 6\alpha(1-\alpha) \de \alpha \ .
\end{equation}
where the last relation is proved in Appendix xx.

Hence Spearman's $\rho$ is a weighted average of $\ell_\alpha$ across $0\le\alpha\le 1$. The weights are $w_\alpha=6\alpha(1-\alpha)$ and integrate to $1$. Spearman's $\rho$ measures overall dependence between $u$ and $v$ while  $\ell_\alpha$ decomposes the overall $\rho$ into local or layer contributions. Note $w_\alpha$ has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. Hence Spearman's $\rho$ places most emphasis on the middle of the distribution which may be undesirable in finance or insurance where tail dependence is critical. Modifying the weights $w_\alpha$ leads to alternate measures of overall dependence further explored in  \sref{saltoverall}.

Layer dependence is a summary measure of a copula.    As any summary measure it can mislead.   However it is not as misleading as for example $\rho$.    Say something about it being a nonparametric summary unlike parameters in a Clayton or Frank copula.  
Mention something about its behaviour for eg Gaussian.

\subsection{Properties of layer--dependence}

Note $\ell_\alpha=1$ if $u=v$,  $\ell_\alpha=-1$ if $u=1-v$ and $u$, $v$ independent implies $\ell_\alpha=0$.   Further $-1\leq \ell_\alpha \leq 1$.   Note also $\ell_\alpha$ is ratio of correlations.

Expanding the covariances  in \eref{definition} and manipulating yields
\begin{equation}\label{gapexp}
\ell_\alpha = \frac{\E(v|u>\alpha)-\E(v|u\leq \alpha)}{\E(u|u>\alpha)-\E(u|u\leq \alpha)}
=2 \left\{\E(v|u>\alpha)-\E(v|u\leq \alpha)\right\}\;,
\end{equation}
where $\E$ calculates expectations. The middle expression in \eref{gapexp} is the expected change in $v$ relative to the expected change in $u$ when $u$ crosses $\alpha$. The latter is $0.5$ for all $\alpha$, yielding the final expression in \eref{gapexp}. Hence large $\ell_\alpha$ implies $v$ is sensitive to movements in $u$ across $\alpha$, indicating strong dependence between $v$ and $u$ at $\alpha$. When $\ell_\alpha=0$, $v$ is unchanged on average when $u$ crosses $\alpha$, and $u$ and $v$ are independent at $\alpha$.

@@

The nine panels in \fref{fillustration} display $(u,v)$ scatterplots of exchangeable copulas $C(u,v)=C(v,u)$, and their layer dependence curves $\ell_\alpha$ for all $0\le\alpha\le 1$. $\ell_\alpha$ and $\alpha$ are plotted on the same axes as $v$ and $u$, respectively. Each copula has Spearman's $\rho=0.6$. \fref{fillustration}  emphasises that copulas with the same overall dependence can a variety of  local dependence structures and that local dependence is  captured with the  layer dependence measure $\ell_\alpha$.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \resizebox{40mm}{!}{\includegraphics{normal.eps}}
            \resizebox{40mm}{!}{\includegraphics{student.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural2.eps}} \\
      \resizebox{40mm}{!}{\includegraphics{gumbel.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural1.eps}}
      \resizebox{40mm}{!}{\includegraphics{clayton.eps}} \\
            \resizebox{40mm}{!}{\includegraphics{frank.eps}}
            \resizebox{40mm}{!}{\includegraphics{structural3.eps}}
      \resizebox{40mm}{!}{\includegraphics{structural4.eps}} \\
    \end{tabular}
    \caption{Copulas all  with the same  $\rho=0.6$ but different layer dependence curves $\ell_\alpha$ over $0\leq\alpha\le 1$ (red curves). The top 3 copulas display relatively strong tail dependence in both lower and upper tail.  The first two copulas in the middle three have relatively strong upper tail dependence and while the third has a high degree of lower tail dependence.   The bottom three copulas have relatively high local dependence in the middle of the distribution and less correlation in the tails.  Overall the panels illustrate how an overall given level of $\rho$ can mask a  range of local dependence: and that level dependence is an appropriate tool for assessing local dependence.}
    \label{fillustration}
  \end{center}
\end{figure}

@@

If a copula is symmetric in $u$ and $v$, then $\ell_\alpha$ is the same if $u$ and $v$ are exchanged.   Further, in this exchangeable case,
 layer dependence $\ell_\alpha$ is closely connected to measures the lack of discordance and dispersion.  In particular
\begin{equation}\label{decomposition}
\ell_\alpha = 1-2(1+\gamma_\alpha) \delta_\alpha \;,
\end{equation}
where
$$
\gamma_\alpha \equiv \cor\{(u\leq\alpha),(v>\alpha)\} =\cor\{(u>\alpha),(v\leq\alpha)\}\;,
$$
$$
\delta_\alpha \equiv\E\left\{(|u-v|)|(u-\alpha)(v-\alpha)<0\right\}  \; .
$$
A proof of \eref{decomposition} is  below.  The correlation $-1\leq\gamma_\alpha\leq 1$ measures the tendency for $(u,v)$ to be discordant at $\alpha$: opposite signs on $u-\alpha$ and $v-\alpha$.  The expectation $0\leq \delta_\alpha\leq 1$ measures the dispersion between discordant points $u$ and $v$ at $\alpha$.

@@

Layer dependence $\ell_\alpha$ can be written as a standardised integral of the copula

\begin{equation}\label{copula}
\ell_\alpha = \frac{\int_0^1 \cov\{I_\alpha(u),I_\beta(v)\} \de\beta }{\alpha(1-\alpha)/2}
= \frac{2 \int_0^1 C(\alpha,\beta) \de \beta- \alpha}{\alpha(1-\alpha)} \;.
\end{equation}
Note $\cov\{(u>\alpha),(v>\beta)\}=C(\alpha,\beta)-\alpha\beta$ and apply \eref{decompose} to $v$, to derive

Thus $\ell_\alpha$ integrates copulas to reduce their dimension from two and one, and scales the result to ensure it lies between $\pm 1$.
